Oriented Gradient Histogram applied to P300 Detection

### 2017-12-13 response.letter.tex

2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. The title of the article was changed accordingly to your suggestion.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.2) there is evidence of the benefits of working with these signals P300, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
You're  right, thank you very much for your comment.  We published in Ramele et al 2016 the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to decode Motor Imagery.  We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and to decode SSVEP patterns.  We added this information on the article (section INTRODUCTION and section CONCLUSION).  We hope this new information enrich the article and clarifies this point.

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\subsection*{\ovalbox{Third Paragraph}}

@@

3.1) Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We worked hard to improve all the mathematical description of the method. Section FEATURE was modified as well as section PARAMETERS.  Section SIGNAL PLOT was revised and its mathematical terminology were corrected.}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

### 2017-12-13 article.tex
Saque finalmente hit y saque el grafico de epoching de los resultados.  Cambie el orden de los pasos, la descripcion del dataset own paso a estar en presente (suena rara), agregue referencias a la expansion de este metodo para otras casos, ajuste el epigrafe de lo que tendria que tener sentido mostrar en un grafico (el grafico por ahora lo deje lo voy a modificar en base al epigrafe) agregue porcentajes en los resultados, puse tambien los resultados hasta ahora con el dataset nuestro, aclare los sujetos de ese dataset, quienes son y que edades tienen, elimine la referencia al zscore, modifique

% Current address and/or shared authorship
\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients.}

% Keywords
\keyword{electroencephalography (EEG); BCI; P300; ALS; NBNN; HOG; SIFT}

@@

%\begin{enumerate}[leftmargin=*,labelsep=3mm]
%\item	Why
%\item	Purpose and significance
%\item	Current state of the research field and key publications cited
%\item   Controversial hypothesis
%\item   Main aim and principal conclusions.
%\end{enumerate}
%
%\begin{enumerate}[leftmargin=*,labelsep=3mm]
%\item El paper tiene un foco general en EEG y rapidamente se centra en BCI como una prueba del método.
%\item La relevancia de estudiar métodos de análisis de EEG para pacientes con ALS.
%\item Ofrece una Receta de cómo implementar un esquema simple de detección de P300 y ofrece una comparativa contra el paper de las italianas.
%\item La transversalidad del método de poder detectar eventos rítmicos asi como transientes como p300.
%\item La posibilidad de utilizarlo para reconocer, fuertemente un patron particular y con la flexibilidad de detectar variantes basado en las poblaciones de descriptores (esta es mi hipótesis de porque funciona, porque yo vi que el método se hace pelota con pequeñas variantes de la señal).
%\item La importancia del averaging de EEG y sus desbalanceos, puntualizando que en el paper original
%\item Que la caída en la clasificación no cae significativamente aún bajando la cantidad de repeticiones requeridas para el averaging.
%\item ¿ Por qué da mejor en PO8 que en Cz y Pz que son los canales P300 ?  (como future work).
%\end{enumerate}

@@

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic patterns using gradient histograms orientations. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events like those produced by P300 Event Related Potential. 
 Its validity is verified by processing offline data for ALS patients.  
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting  raw signal plots~\citep{Hartman2005}.

@@

%\subsection{Morphological Signal Analysis}
%
%Several approaches and methods have been applied to decode them.  Of particular interests are morphological signal analysis 
%CITE (Paper del chabon del ITBA que detalla bien los diferentes metodos) \citep{Alvarado-Gonzalez2016,Yamaguchi2009}.
%
%Morphological or 
%
%template matching
%
%spike
%sharp wave
%spike wave
%polyspike 
%slow wave complex
%
%shape domain
%
%\subsection{P300}
%
%Uhhhh I can talk a lot about P300 \citep{Knuth2006}.
%
%P300 single trial is a golden grail of Brain computer interfaces.  It is studied and analyzed in this way, here in this other way, in this other way is here.

@@

%The proposed algorithm was designed to perform specifically this decoding.

% of the selected letter based on the detection of the P300 Event Related Potential response on the corresponding EEG signal.

\subsubsection{Preprocessing} \label{Preprocessing}

The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. 
The processing pipeline starts by applying a notch filter to the raw signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.

@@

\begin{itemize}
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments are labeled as \textit{hit}, whereas the remaining $10$ are labeled as \textit{no hit}.  A hit represents that the EEG segment should contain the P300 ERP signature time-locked to the flashing stimulus. 
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

The underlying idea of this method is to generate a template for the signal \citep{Ramele2016},  based on the image plot.  Hence, the first step is the signal transformation into a temporary binary image.

Averaged signal segments obtained from the preprocessing stage are first scaled and standardized (i.e. z-score) by

%\noindent and is used to derive local representations that captures the visual shape of the signal.

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, descriptors from \textit{hit} labeled segments are extracted.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The process has the following steps:

@@

%In the case of the P300 response, the oddball paradigm requires that one of the stimuli be infrequent. Hence this forces the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}. %Para solucionar este problema,

@@

%------------------
\subsection{Experimental Protocol} \label{Protocol}

@@

We replicated the same experiment on healthy subjects using a wireless digital EEG device (g.Nautilus, g.Tec, Austria).  The experimental conditions were the same as those used for the previous dataset, as detailed in section~\ref{ALSDataset}.

Participants were recruited voluntarily and the experiment were conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation was handed out and all participants agreed and signed a written informed consent.  All healthy subjects had normal or corrected-to normal vision and no history of neurological disorders. The experiment was performed with 20 subjects, 7 males, 4 females, average age 22.5 years, standard deviation  1,8 years, range 21–26 years.

Gel-based active electrodes (g.LADYbird, g.Tec, Austria) were used on the same locations Fz, Cz, P3, Pz, P4, PO7, PO8 and Oz.  Reference was set to the right ear lobe and ground was preset as the AFz position.   Sampling frequency is slightly different, and was set to 250 Hz, which is the closest possible to the one used with the other dataset.

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ in the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds in the horizontal direction. Finally, descriptors  $\mathbf{kp}$  are located at $ x_{kp} =  0.55 Fs.\gamma = 35 $ and $ y_{kp} = z(c) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction). 
% esta figura no está comentada en el texto.
\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole signal information}
\label{fig:sampledescriptor2}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion} \label{Results}
\label{section:results}
In Figure \ref{fig:subjectaveraged} the grand average (point-to-point) for all the subjects using the information from all the segments can be shown.  The P300 characteristic curve can be seen particularly in subjects 2, and 6 and in a lesser extend in the remaining subjects. In order to correctly decode the selected letter from each trial, particular care was observed to avoid unbalanced number of epochs  (i.e. an unequal number of epochs on each condition), because that may introduce bias in the classification procedure (the variance of averaged signals is inversely proportional to the number of samples and the procedure would be discriminating signals with different variances).

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{subjectaveraged.eps}
\caption{Point-to-point grand averages of epochs obtained for hits (solid line) and no hits (dashed line) for each one of the 8 subjects for channel Cz. The P300 characteristic curve can be well identified particularly on subjects 2 and 6.}
\label{fig:subjectaveraged}
\end{figure}

Results are shown in Table \ref{tab:results} where the percentage of correctly spelled letters is calculated while performing an offline BCI Simulation.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well.

DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel.  However, in our approach

It is of particular interest that using this method, the best performing channel was not always Cz, and instead occipital channels PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject. Chance level is 0.02 }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   0.35  \\
2     &     Fz   &   0.85  \\
3     &     Cz   &   0.25  \\
4     &     PO8 &   0.55  \\
5     &     PO7 &   0.40 \\
6     &     PO7 &   0.60  \\
7     &     PO8 &   0.80  \\
8     &     PO7 &   0.95 \\

@@

The ITR, or BTR, in the case of reactive BCIs \citep{WolpawJonathanR2012} strongly depends on the amount of signal averaging required to transmit a valid and robust selection.  The Performance curves (Fig. \ref{fig:performance}) show how the percentage of correctly identified letters depends on the number of repetitions that were used to obtain the averaged signal.

@@

We found that only by using the visual aspect of the P300 signal plot,  for some subjects it was not possible to find templates that could allow the classification to reach a higher level.  We hypothesize that as subjects may have different latencies and amplitudes of the P300 signal \citep{Riccio2013}, it may also be the case that the shape of the generated ERP may vary greatly in an intra-subject manner, thus the pattern could not be well generalized and a higher performance could not be reached (Fig.  \ref{fig:p300templates}).

@@

@@

We believe that the expanding and the understanding of this tool in order to automatically classify those patterns in EEG that are specifically identified by their shapes (e.g. K-Complex, Vertex Waves, Positive Occipital Sharp Transient \citep{Hartman2005}) is a prospect future work to be considered.  It may also provide  assistance to physician or electroencephalographers to help them locate these EEG patterns particularly in long recording periods, frequent in sleep research.

>>>>

% Current address and/or shared authorship  COMMENTED PER REVIEWER
%\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

\firstnote{.}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{EL OBJETIVO DE ESTE PAPER, EL HILO CONDUCTOR, ES DEMOSTRAR QUE ESTE METODO QUE ESTA BASADO EN LA ESTRUCTURA DE LA SENIAL SE PUEDE UTILIZAR TAMBIEN PARA IDENTIFICAR EVENTOS TRANSIENTES COMO EL P300, Y QUE CON ESO SE PUEDE IMPLEMENTAR UN SPELLER DE P300.  The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset for healthy subjects.}

% Keywords
\keyword{EEG; BCI; P300; ALS; NBNN; HOG; SIFT}

@@

@@

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves~\citep{Ben-Simon2013}  and Motor Imagery rolandic central $\mu$ rhythms~\citep{Hartman2005} using gradient histograms orientations. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  Its validity is verified by processing offline data for ALS patients and for healthy subjects. 
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots~\citep{Hartman2005}.

@@

@@

The underlying idea of this method is to generate templates for the signal \citep{Ramele2016}, based on the image plot.  Hence, the first step is the signal transformation into a temporary binary image.

@@

\begin{itemize}
\item \textbf{Preprocessing}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, two descriptors are extracted from averaged signal segments which correspond to the letter where the user was supposed to be focusing onto.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The process has the following steps:

@@

@@

%------------------
%In the case of the P300 response, the oddball paradigm requires that one of the stimuli be infrequent. Hence this forces the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}. %Para solucionar este problema,

Averaged signal segments are scaled and standardized by

\subsection{Experimental Protocol} \label{Protocol}

@@

We replicate the same experiment on healthy subjects using a wireless digital EEG device (g.Nautilus, g.Tec, Austria).  The experimental conditions are the same as those used for the previous dataset, as detailed in section~\ref{ALSDataset}.

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 20 subjects, 17 males, 3 females, average age 26.85 years, standard deviation  8.8 years, range 19-50 years.

Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on the same locations Fz, Cz, P3, Pz, P4, PO7, PO8 and Oz.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is slightly different, and is set to 250 Hz, which is the closest possible to the one used with the other dataset.

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ in the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds in the horizontal direction. Finally, descriptors  $\mathbf{kp}$  are located at $ x_{kp} =  0.55 Fs.\gamma = 35 $ and $ y_{kp} = z(c) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction). 
% esta figura no está comentada en el texto.
\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole signal information.  The vertical size $4 \cdot 3 s_y$ of the patch is required to be long enough as to contain the signal $\Delta  \mu V $, which is the peak-to-peak amplitude of the transient event. The horizontal size $ 4 \cdot 3 s_x$ , on the other hand, needs to cover the entire duration in seconds of the transient signal event, $ \lambda $.   }
\label{fig:sampledescriptor2}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion} \label{Results}
\label{section:results}

%In Figure \ref{fig:subjectaveraged} the grand average (point-to-point) for all the subjects using the information from all the segments can be shown.  The P300 characteristic curve can be seen particularly in subjects 2, and 6 and in a lesser extend in the remaining subjects. In order to correctly decode the selected letter from each trial, particular care was observed to avoid unbalanced number of epochs  (i.e. an unequal number of epochs on each condition), because that may introduce bias in the classification procedure (the variance of averaged signals is inversely proportional to the number of samples and the procedure would be discriminating signals with different variances).  
%
%\begin{figure}[H]
%\centering
%\includegraphics[width=16cm]{subjectaveraged.eps}
%\caption{Point-to-point grand averages of epochs obtained for hits (solid line) and no hits (dashed line) for each one of the 8 subjects for channel Cz. The P300 characteristic curve can be well identified particularly on subjects 2 and 6.}
%\label{fig:subjectaveraged}
%\end{figure}

Results are shown in Table \ref{tab:results} where the percentage of correctly spelled letters is calculated while performing an offline BCI Simulation for the public dataset of ALS patients.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well.

DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel, more representative of the difference in amplitudes. However, in our approach occipital channels,  PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject. Chance level is $2\%$. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   $35\%$  \\
2     &     Fz   &   $85\%$  \\
3     &     Cz   &   $25\%$  \\
4     &     PO8 &   $55\%$  \\
5     &     PO7 &   $40\%$ \\
6     &     PO7 &   $60\%$  \\
7     &     PO8 &   $80\%$  \\
8     &     PO7 &   $95\%$ \\

@@

The ITR, or BTR, in the case of reactive BCIs \citep{WolpawJonathanR2012} strongly depends on the amount of signal averaging required to transmit a valid and robust selection.  The Performance curves (Fig. \ref{fig:performance}) show how the percentage of correctly identified letters depends on the number of repetitions that were used to obtain the averaged signal.
For both datasets, the experimental protocol uses a very short ISI which has the potential to increase the ITR but at the same time is inversily proportional to the amplitude of the p300 response (i.e. harder to detect).  As an unexpected consequence, 15 out of 20 subject reported high ocular tiredness by using this configuration.

@@

We found that only by using the visual aspect of the P300 signal plot,  for some subjects it was not possible to find templates that could allow the classification to reach a higher level.  Moreover, as can be seen in the figure, as repetitions tend to 1, which will correspond to single trial letter identification, the performance suffers greatly as the signal plot is simple too complex to be used as template and the method is completely unable to find good matches  PODRIA ACA MOSTRAR ADEMAS PLOTS DE SINGLE TRIAL PARA QUE VEAN QUE SON UN QUILOMBO.

UNA DE LAS RAZONES POR LA CUAL PODEMOS LLEGAR A ARGUMENTAR QUE NO FUIMOS MAS A FONDO CON EL TEMA ES QUE NO QUERIAMOS DISTORSIONAR DEMASIADO LA SENAL COMPLEJIZANDO EL ALGORITMO Y QUERIAMOS MANTENERNOS ENTRE LO MAS ESTANDAR POSIBLE (PUEDO CITAR MAS O MENOS QUE NUESTRO PROCESSING PIPELINE ES ESTANDARD).

We hypothesize that as subjects may have different latencies and amplitudes of the P300 signal \citep{Riccio2013}, it may also be the case that the shape of the generated ERP may vary greatly in an intra-subject manner, thus the pattern could not be well generalized and a higher performance could not be reached (Fig.  \ref{fig:p300templates}).  It is known (REFERENCE) that the point-to-point averaging suffers from inter-trial variability (REFERENCIA A ESE PAPER Con ese nombre) due to latencies shifts.  We tested a naive approach to verify if the morphological shape of the p300 can be stabilized by applying different shifts and we verified that there is a better performance when a correct resynchronization is applied.  We also applied DTW but couldn't found a sustantial improvement.

Additionally we believe that our method suffers also from null signals.

ACA EL PUNTO PARA MOSTRAR ES QUE DE ALGUNA MANERA ESTAMOS COMO MOSTRANDO UN "SI SE PUEDE" DE UN TRABAJO QUE PUEDE CONTINUAR Y EXPANDIRSE.   POR ESO LA IDEA DE MOSTRAR PUNTAS DE COSAS QUE VIMOS E HICIMOS, PERO SIN MOSTRAR RESULTADOS CONCRETOS.

Considering the amplitude variation of the p300 and its components, we proposed a conservative approach by standarizing the signal (which can have both effects, of reducing noise that were not reduced by the averaging procedure but at the same time it can reduce the amplitude of a significant p300 compoent).   The standarization also provides another benefit which is sometimes neglected as unimportant.  The variance of the averaged signal depends on the number of signals averaged.  As this method analyze precisely the shape of the signal in the plot, which is very affected by the variance, the method will not be able to discriminate solely based on the underlying cognitive event, it will be confused by the number of signals that were used to calculate the averaged signal.

The SIFT method is invariant to the scale, but in order to allow our method to be also invariant to the amplitude of the signal we had to modify the geometry of the descriptor to remove the squared constrain.

@@

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each healthy subject. Chance level is $2\%$. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &     0.7000  &  0.3500  \\
1     &     Cz   &    0.1000  &  0.0500 \\
1     &     Cz   &     0.1000  &  0.1000 \\
1     &     Cz   &     0.1000  &  0.1500 \\
1     &     Cz   &     0.1000  &  0.1000 \\
1     &     Cz   &     0.5500  &  0.4000 \\
1     &     Cz   &     0.1500   & 0.1000 \\
1     &     Cz   &     0.4500   & 0.2000 \\
1     &     Cz   &     0.3000  &  0.2000 \\
1     &     Cz   &     0.2500  &  0.1500 \\
1     &     Cz   &     0.5500  &  0.3000 \\
1     &     Cz   &     0.3500  &  0.1000 \\
1     &     Cz   &     0.3000  &  0.3500 \\
1     &     Cz   &     0.5000  &  0.4000 \\
1     &     Cz   &     0.3000  &  0.1500 \\
1     &     Cz   &     0.3000  &  0.3000 \\
1     &     Cz   &     0.3500  &  0.3500 \\
1     &     Cz   &          0      & 0.0500 \\
1     &     Cz   &     0.7000   & 0.5000 \\
1     &     Cz   &     0.4000  &  0.3000 \\

\bottomrule
\end{tabular}
\label{tab:resultsowndataset}
\end{table}

In Table \ref{tab:resultsowndataset} results obtained for the 20 healthy subjects is shown.  The third column shows the results obtained by using the Krusienski feature and the SVM classifier.  On the other hand, the fourth column shows the results obtained by the present method as described in previous sections.  Subject 17 was used as control as he was instructed to count only the letters and to stare at the screen but not neglected row/column flashing.

@@

We believe that the expanding and the understanding of this tool in order to automatically classify those patterns in EEG that are specifically identified by their shapes (e.g. K-Complex, Vertex Waves, Positive Occipital Sharp Transient \citep{Hartman2005}) is a prospect future work to be considered. We are currently working in unpublished material analyzing KComplex that may also provide  assistance to physician or electroencephalographers to help them locate these EEG patterns particularly in long recording periods, frequent in sleep research.
